A Classical Density-Functional Theory for Describing Water Interfaces

### 2012-06-01 papers/water-SAFT/paper.tex
Reran canculations for four-rods 1.8nm in diameter. Moved contact value section of paper (still needs work). Changed plot for pressure vs. density and added more experimental data. Tested everything on quipu.

A number of classical density functionals have been developed for
water~\cite{ding1987, Yang1992, gloor2002saft, Jaqaman2004,
  clark2006developing, lischner2010classical}\fixme{ADD REST OF
  LIST HERE!}, capturing some of the qualitative behavior of water.
However, many of these are less sophisticated than those describing
the hard-sphere fluid, and a quantitatively accurate functional
describing water---which reproduces the temperature dependence of the
macroscopic surface tension, its decrease as curvature increases, and
its quantitative value---has not yet been published.  For instance,
the functional of Lischner \emph{et al}\cite{lischner2010classical}
treats the surface tension correctly, but can only be used at room
temperature.  A numer of classical density functionals have recently
been produced that are based on Statistical Associating Fluid Theory
(SAFT)\cite{gloor2002saft, clark2006developing}\fixme{add
  more}.  These functionals are based on a perturbative thermodynamic
expansion, and give a reasonable prediction for the temperature
dependence of the surface tension.  However, they also overestimate
the surface tension by a factor of about 1.5, and are unsuited for studying
systems with molecular-scale inhomogeneities.

\subsection{Contact-value theorem}

The contact value theorem states that the pressure of a fluid at a
hard wall interface is proportional to the contact density of the fluid,
$n_c$.  Specifically,
\begin{equation}\label{eq:contact}
  P = n_ck_BT
\end{equation}
where $k_B$ is the Boltzmann constant and $T$ is the temperature of the
fluid. This leads to the property that the free energy of solvation of
a hard solute is proportional to its volume in the limit of small
solutes: 
\begin{equation}\label{contactvaluethm}
  F = n_c k_BT V\:
\end{equation}

@@

The SAFT Helmholtz free energy is composed of five terms:
\begin{equation}
  F[n] = F_\textit{id} + F_\textit{hs} + F_\textit{disp} +
  F_\textit{assoc} + F_\textit{chain},
\end{equation}
where the first three terms encompass the \emph{monomer} contribution
to the free energy, the fourth is the \emph{association} free energy,
describing hydrogen bonds, and the final term is the chain formation
energy of for fluids that are chain polymers.  The final term is not
relevant for water.  While a number of SAFT formulations have been
published, we will focus on SAFT-VR\cite{gil-villegas-1997-SAFT-VR},
which has been used by Clark \emph{et al} to construct an optimal SAFT
model for water\cite{clark2006developing}.

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{figs/pressure-with-isotherms}
\end{center}
\caption{The pressure versus density for various temperatures, including
experimental pressure data from NIST\cite{nistwater}. The dash-dotted lines
indicate the computationally calculated pressure and the open circles are 
NIST data points, plotted for every other isotherm. The solid and dotted lines
represent the theoretical and experimental coexistence curves.}
\label{fig:pressure-with-isotherms}
\end{figure}

SAFT has also been used to construct classical density functionals,
which have been applied to study the surface tension as a function of
temperature\cite{whotocite}.  Such functionals have qualitatively
predicted the dependence of surface tension on temperature, but have
in general provided poor quantitative agreement with experiment.
Moreover, the vast majority of SAFT-based classical density
functionals have made local-density approximations that make them
unsuitable for application to systems that are inhomogeneous at the
molecular length scale\cite{felipe2001examination, gloor2002saft,
  gloor2004accurate, clark2006developing, gloor2007prediction,
  kahl2008modified, gross2009density}.

\section{Theory and Methods}
We construct a new classical density functional for water. We treat
each individual water molecule as a purely classical hard sphere and
add association terms based on SAFT. For our free energy, we use the
first four terms from equation (number): $F_\textit{id}$,
$F_\textit{hs}$, $F_\textit{disp}$ and $F_\textit{assoc}$. In
addition, a chemical potential term is used, so we actually work in
the grand canonical ensemble.  In the following sections, we will
introduce the terms of this functional.

>>>>

A number of classical density functionals have been developed for
water~\cite{ding1987, Yang1992, gloor2002saft,gloor2004accurate,
  gloor2007prediction, Jaqaman2004, clark2006developing,
  lischner2010classical, fu2005vapor-liquid-dft,felipe2001examination}
, capturing some of the qualitative behavior of water.
However, many of these are less sophisticated than those describing
the hard-sphere fluid, and a quantitatively accurate functional
describing water---which reproduces the temperature dependence of the
macroscopic surface tension, its decrease as curvature increases, and
its quantitative value---has not yet been published.  For instance,
the functional of Lischner \emph{et al}\cite{lischner2010classical}
treats the surface tension correctly, but can only be used at room
temperature.  A number of classical density functionals have recently
been produced that are based on Statistical Associating Fluid Theory
(SAFT)\cite{gloor2002saft, clark2006developing, gloor2007prediction,
  gloor2004accurate, gross2009density, kahl2008modified, felipe2001examination}.
These functionals are based on a perturbative thermodynamic
expansion, and give a reasonable prediction for the temperature
dependence of the surface tension.  However, they also overestimate
the surface tension by a factor of about 1.5, and \fixme{most?} are unsuited
for studying systems that have density variations on a molecular
length scale due to the use of a local density approximation.

@@

The SAFT Helmholtz free energy is composed of five terms:
\begin{equation} \label{eq:SAFT-free-energy}
  F[n] = F_\textit{id} + F_\textit{hs} + F_\textit{disp} +
  F_\textit{assoc} + F_\textit{chain},
\end{equation}
where the first three terms encompass the \emph{monomer} contribution
to the free energy, the fourth is the \emph{association} free energy,
describing hydrogen bonds, and the final term is the chain formation
energy of for fluids that are chain polymers.  The final term is not
relevant for water.  While a number of SAFT formulations have been
published, we will focus on SAFT-VR\cite{gil-villegas-1997-SAFT-VR},
which has been used by Clark \emph{et al} to construct an optimal SAFT
model for water\cite{clark2006developing}.

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{figs/pressure-with-isotherms-truncated}
\end{center}
\caption{The pressure versus density for various temperatures, including
experimental pressure data from NIST\cite{nistwater}. The dash-dotted lines
indicate the computationally calculated pressure and the open circles are 
NIST data points. The solid and dotted lines
represent the theoretical and experimental coexistence curves.}
\label{fig:pressure-with-isotherms}
\end{figure}

SAFT has also been used to construct classical density functionals,
which have been applied to study the surface tension as a function of
temperature\cite{whotocite}.  Such functionals have qualitatively
predicted the dependence of surface tension on temperature, but have
in general provided poor quantitative agreement with experiment.
Moreover, the vast majority of SAFT-based classical density
functionals have made local-density approximations that make them
unsuitable for application to systems that obey the contact-value
theorem. \cite{felipe2001examination, gloor2002saft,
  gloor2004accurate, clark2006developing, gloor2007prediction,
  kahl2008modified, gross2009density}.\

 However, they also overestimate
the surface tension by a factor of about 1.5, and \fixme{most?} are unsuited
for studying systems that have density variations on a molecular
length scale due to the use of a local density approximation.

Functionals constructed using a local density approximation
  will always fail to satisfy the contact-value theorem, and will
  therefore incorrectly model small hard solutes.
The contact-value theorem states that the pressure of any fluid at a
hard wall interface is proportional to the contact density of the
fluid\cite{henderson1979exact}, according to the following relation:
\begin{equation}\label{eq:contact}
  p = n_ck_BT
\end{equation}
where $n_c$ is the contact density, $k_B$ is the Boltzmann constant
and $T$ is the temperature of
the fluid. This leads to the property that the free energy of
solvation of a small hard solute is proportional to the
solvent-excluded volume:
\begin{equation}\label{contactvaluethm}
  F = n k_BT V
\end{equation}
\fixme{add this here? :} The contact-value theorem is satisfied for
functionals for with the only purely local term is the free energy of
an ideal gas.

\section{Theory and Methods}
We construct a new classical density functional for water. We treat
each individual water molecule as a purely classical hard sphere and
add association terms based on SAFT. For our free energy, we use the
first four terms from Equation \ref{eq:SAFT-free-energy}: $F_\textit{id}$,
$F_\textit{hs}$, $F_\textit{disp}$ and $F_\textit{assoc}$. In
addition, a chemical potential term is used, so we actually work in
the grand canonical ensemble.  In the following sections, we will
introduce the terms of this functional.
